\documentclass[11pt]{article}
\usepackage{RCS,verbatimstyle}
\SVN    $Id: prop.tex 72 2012-07-04 22:10:41Z sufrin $
%
%
%
\usepackage[bottom]{footmisc}
\usepackage[mathletters]{ucs}
\usepackage[utf8x]{inputenc}
%
% Sanserif blackboard bold
%
\RequirePackage[sans]{dsfont}
\def\mathbb#1{\mathds{#1}}
%
%
%
\usepackage{japecode}
\usepackage[colorlinks,linkcolor=blue,filecolor=blue]{hyperref} 
%
%
%
\title{Propositional Logic and Proof \\\textit{via}\\ Functional Programming}
\author{Bernard Sufrin\thanks{Oxford University Computer Science Department}}
\date{Hilary Term, 2013}
%
%
%
%\codestyle{haskell}
\parindent=0pt\parskip=\medskipamount
%
%
%
\begin{document}
\maketitle
\begin{abstract}
\noindent In these notes we give an account of the
development of a simple \textit{proof assistant} for
propositional logic. Our intention is to demonstrate a useful way of
thinking about propositions and proofs as data structures.
\end{abstract}

\tableofcontents

\newpage
\section{Introduction}

\section{Propositions}

\subsection{Representation}
We represent propositions as recursive data type. The
comment by each constructor is an example of the notation
used to write that kind of proposition.

\begin{code}[propositions.hs]
module Propositions where
  
  ... Import types classes and functions from the Prelude

  data Prop = Atom Atomic         -- a, b ...
            | Not  Prop           -- not p
            | Prop `And` Prop     -- p ∧ q
            | Prop `Or`  Prop     -- p ∨ q
            | Prop `Imp` Prop     -- p ⇒ q
            | Prop `Iff` Prop     -- p ⇔ q
            deriving (Eq)
\end{code}

\subsection{Haskell notation}
We want to make it convenient to write propositions in Haskell itself; so we 
define Haskell operator symbols corresponding to appropriate
constructors of the \textsf{Prop} type, and give them conventional
syntactic properties.\footnote{It is unfortunate that
we can do this only for infix operators, for we would
have liked to use $\lnot$ as a negation symbol; but there is
no way of defining a \textit{symbolic} prefix operator in Haskell.} 

\begin{code}[]
  infixl 9 ∧;  (∧)  = And
  infixl 8 ∨;  (∨)  = Or
  infixr 7 ⇒; (⇒)  = Imp
  infixl 6 ⇔; (⇔)  = Iff  
  not = Not
\end{code}

We are not interested in the internal structure of atomic
propositions, but it will be convenient to represent
them as single letters. Here we invent Haskell names for a few atomic
propositions.
\begin{code}
  type Atomic = String
  [a,b,c,d,p,q,r,s] = map (Atom . (:"")) "abcdpqrs"
\end{code}


\subsection{Printing}
We want propositions to be printed using the same notation
that we write them in (in Haskell). We declare \textsf{Prop}
to be an instance of the \textbf{Show} class, and define a
function \textsf{showProp} that implements
\textsf{showsPrec} for propositions.

\begin{code}[]
  instance (Show Prop) where showsPrec = showProp  
  showProp::  Int -> Prop -> ShowS
  showProp d (Atom a)    = showString a
  showProp d (Not  p)    = showString "not " . showProp 10 p
  showProp d (p `And` q) = showInfix d 9 9 " ∧ "     p q
  showProp d (p `Or` q)  = showInfix d 8 8 " ∨ "     p q
  showProp d (p `Imp` q) = showInfix d 7 6 " ⇒ "    p q
  showProp d (p `Iff` q) = showInfix d 5 5 " ⇔ "    p q
\end{code}

The auxiliary function \textsf{showInfix} handles the
showing of binary composite propositions, and takes account
of the right-associativity of implication in the usual way,
namely by giving implication a slightly higher syntactic
priority on its right than on its left. 

\begin{code}
  showInfix:: Int -> Int -> Int -> String -> Prop -> Prop -> ShowS
  showInfix d d' d'' op l r  =  
            showParen (d>=d') $ 
            showProp  d' l   .
            showString op    . 
            showProp  d'' r
\end{code}

Here are some examples of propositions

\begin{code}[]
  p0 = (a ⇒ b) ⇒ c
  p1 = a ⇒ b ⇒ c
  p2 = a ∧ b ⇒ c ∧ d
  p3 = not a  ∧  b  ⇒  not c  ∧  d
  p4 = (a ⇒ (b ⇒ c)) ⇒ ((a ⇒ b) ⇒ c)
  p5 = not p4
\end{code}

and here's what the Haskell interpreter prints for them. Notice that the redundant
parentheses on the right of $p4$ are omitted, but that in all other respects 
the Haskell output was identical to the Haskell input.
\begin{verbatim}
  > [p0, p1, p2, p3, p4, p5]
  [       (a ⇒ b) ⇒ c
  ,       a ⇒ b ⇒ c
  ,       a ∧ b ⇒ c ∧ d
  ,       not a ∧ b ⇒ not c ∧ d
  ,       (a ⇒ b ⇒ c) ⇒ (a ⇒ b) ⇒ c
  ,       not ((a ⇒ b ⇒ c) ⇒ (a ⇒ b) ⇒ c)
  ]
\end{verbatim}


\section{Proof Representation}
\begin{code}
  newtype Sequent = [Prop] :|- Prop
  infix 1 |-
  infix 1 :|-
  hyps |- conc = hyps :|- conc
      
  instance (Show Sequent) where 
     show (hyps :|- conc) = showHyps hyps ++ " :|- " ++ show conc
     
  data    Proof   = Proven [Proof] Sequent Rule
                  | Gap
                  
  instance (Show Proof) where
     showsPrec = showProof 
     
  showHyps []     = ""
  showHyps [h]    = show h
  showHyps (h:hs) = show h ++  ", " ++ showHyps hs
  
  showProof :: Int -> Proof -> ShowS
  showProof level Gap = indentBy level . showString "..."
  showProof level (Proven subproofs seq rule) =
            foldAll showSubproof subproofs .
            indentBy level                 .
            showsPrec 0 seq                .
            showString " by "              .
            showString rule
            where 
            showSubproof proof s = showProof (level+1) proof ("\n" ++ s)
            foldAll f xs ys = foldr f ys xs         
                  
  indentBy:: Int -> ShowS
  indentBy level s = take (2*level) spaces ++ s
  
  spaces:: String
  spaces = ' ':spaces 
  
  type Rule = String

  goal1 = [p ⇒ (q ⇒ r)] :|- p ⇒ r

  
\end{code}

\section{Proof Rules}
\begin{code}[]
{-
        RULE  "⇒⊢"
        FROM  Γ       ⊢ A
        AND   Γ, B    ⊢ C
        INFER Γ, A⇒B ⊢ C 
        
        RULE  "⊢⇒"
        FROM  Γ, A ⊢ B
        INFER Γ    ⊢ A⇒B
-}
\end{code}
\begin{code}[]
  (∪) = List.union
  (∩) = List.intersect
  (ε)  = List.member
  
  
\end{code}

\section{Related Work}

\href{http://pauillac.inria.fr/~leifer/articles/logic/LogicPearl.dvi.gz}
{Deduction for functional programmers} by James J. Leifer and Bernard Sufrin.
Journal of Functional Programming, volume 6, number 2, 1996.

This paper was the result of research we did together while James Leifer was an undergraduate 
in Oxford. It describes a new way of
teaching formal deductive logic to students familiar with functional
programming and introduces to them metamathematical ideas by
exploiting the students' familiarity with freely-generated
data-structures and their recursion and induction principles.

\href{http://pauillac.inria.fr/~leifer/articles/logic/LogicviaFP.300.ps.gz}
{Formal logic via functional programming} by James J. Leifer (June 1995).
Final year dissertation, Oxford University, supervised by
C.A.R. Hoare and Bernard Sufrin.

This report is a greatly expanded version of the above paper. See
the
\href{http://pauillac.inria.fr/~leifer/articles/logic/LogicviaFP.Errata}
{Errata} list, the Gofer\footnote{Haskell's immediate predecessor}
\href{http://pauillac.inria.fr/~leifer/articles/logic/LogicviaFP.gs}
{source-listings} extracted from the document, and the
\href{http://pauillac.inria.fr/~leifer/articles/logic/LogicviaFP.j} {JAPE
source-listings} (which are no longer compatible with the
latest version of JAPE).

\appendix
\section{Deferred sections of code}

\subsection{... import types classes and functions from the Prelude}
\begin{code}[... import types classes and functions from the Prelude]

  import Prelude(Eq, String, Show, ShowS, Int, 
                 map, (.), (++), foldr, take, unlines, (>=), (+), (*), ($), 
                 showParen, showString, showsPrec, show)
  import List

\end{code}

\begin{code}[]
 -- A supplementary deferred section
 
\end{code}

\begin{code}[propositions.hs]
 -- A final comment
 
\end{code}
\end{document}






